A dynamic bioenergetic model for coral-\emph{Symbiodinium} symbioses and coral bleaching as an alternate stable state

### 2017-07-14 ms/ms.Rmd
text edits/corrections

1. _Develop a new module of photo-oxidative stress._ Of primary interest to coral biologists and ecologists is symbiosis dysfunction under environmental stress, resulting in coral “bleaching”--the loss of algal symbionts from the association [@Jokiel:1977p7353]. Bleaching is thought to be triggered by photo-oxidative stress in *Symbiodinium* [@Weis:2008p944], which has been modeled previously as a response to absolute external irradiance [@Eynaud:2011tv]. However, this response may also depend on self-shading by other symbionts [@Enriquez:2005p142], CO~2~ availability at the site of photosynthesis [@Wooldridge:2009p7807], and other non-photochemical quenching mechanisms [@Roth:2014wf]. We incorporate these features into a new photo-oxidative stress module linking overreduction of the photosynthetic light reactions to downstream impacts of photoinhibition and photodamage. Importantly, this formulation introduces a link between CO~2~-limitation of photosynthesis and bleaching, and potential synergistic roles for heterotrophy and nutrient availability in influencing bleaching responses [@Wooldridge:2014hc]

@@

3. _Provide well-documented, open source code._ In order to facilitate the continued development and application of theoretical modeling tools for coral symbioses, we provide access to the model in the form of an R package called coRal [github.com/jrcunning/coRal](github.com/jrcunning/coRal), which users may install to run and visualize model simulations. With an accessible and modular framework, we envision this code base as a resource for further development by the scientific community to include additional complexity and problem-specific components. We chose R [@RALanguageandEn:2014wf], an open source programming language in common use by biologists and ecologists, to reach the widest possible audience with this work.

@@

Coral biomass formation is then specified by a parallel complementary SU (formula in Fig. 3.7 of @Kooijman:2010vd]). The general form of this equation is $(m^{-1} + x^{-1} + y^{-1} - (x + y)^{-1})^{-1}$, where $m$ is the maximum production flux, and $x$ and $y$ are the input fluxes of the two substrates. Coral biomass is thus produced from carbon and nitrogen according to:

@@

where $\rho_C$ is fixed carbon shared by the symbiont (see Eq. 21), and $r_{NH}$ is recycled nitrogen liberated by host biomass turnover (see Eq. 7). The parameter $y_C$ specifies the yield of biomass from organic carbon, which we take to be 0.8 to satisfy redox balance (see @Muller:2009io).

@@

where $j_{CPm}$ is the maximum specific rate of photosynthesis, and $c_{ROS}$ is the relative rate of reactive oxygen species production (see Eq. 18). Dividing the photosynthetic rate by $c_{ROS}$ causes a decline in response to photooxidative stress at high light levels, and the emergent outcome of this SU formulation demonstrates a classic photoinhibition response (Fig. S2).

@@

The model dynamics are specified by the differential equations (1-2) that impose biomass balance for host and symbiont and by a set of coupled non-linear algebraic equations (3-22) that define fluxes. Several of these fluxes are defined *implicitly*; for example, the rejection fluxes of carbon and nitrogen from the symbiont and host biomass SUs, respectively, act as reciprocal input fluxes to the other SU. Similarly, the photosynthesis SU receives CO~2~ at a rate proportional to the carbon rejection flux from the host biomass SU, and the rejection flux of excitation energy from the photosynthesis SU acts to reduce its own production through photoinhibition. Without further assumptions, however, the dynamical system is not always unambiguously defined because for some combinations of parameters and environmental forcing functions the system of algebraic equations has more than one solution with all fluxes non-negative (see results below). In such circumstances, the right hand side of the differential equations (1) and (2) is not uniquely defined even when $S$ and $H$ are specified. We resolved this problem by *defining the dynamical system* as the limit as a time step $\Delta t \rightarrow 0$ of a discretized system corresponding to Euler integration of the differential equations, with those fluxes that represent flows of elemental matter implemented by assuming that transfer of material between components of the system takes one time step. Thus, for example, CO~2~ rejected from the host SU at time $t$ arrives at the photosynthesis SU at time $t + \Delta t$.

Simulations using the discretized scheme were performed using R code developed in the coRal R package [github.com/jrcunning/coRal](github.com/jrcunning/coRal). By experimentation, we found that a time step of 0.1 days gave adequate precision for most simulations (including used to generate Figs. 2-8 in this paper). For steady state estimations, simulations were run until the changes in specific growth rate of the host and the S:H biomass ratio were less than 1e-5 per time step. In regions of state space where very slow transient dynamics could be expected (i.e. near bifurcation points), sample steady state calculations were verified using MATHEMATICA code for numerical root finding (function FindRoot) with the code written independently by a coauthor without reference to the R code. All of the R code for the simulations and figures presented in this paper can be found in the accompanying data repository at [github.com/jrcunning/coRal-analysis](github.com/jrcunning/coRal-analysis).

@@

At irradiance levels above ~25 mol photons m^-2^ d^-1^, steady state growth rates decline until positive growth ceases above ~40 µmol photons m^-2^ d^-1^ (Fig. 3A). The mechanism underlying this decline is the increase in light energy beyond the capacities of photosynthesis and non-photochemical quenching: excess excitation energy generates reactive oxygen species (ROS) [@Weis:2008p944; @Roth:2014wf], which, in this model, have the phenomenological consequences of reducing the photosynthetic rate (representing photoinhibition) and increasing symbiont biomass loss (representing photodamage and/or symbiont expulsion) [see @Eynaud:2011tv]. Together, these impacts reduce the symbiont to host biomass ratio (Fig. 3B), as occurs during coral bleaching. This reduction in symbionts consequently reduces the flux of fixed carbon to the host, resulting in increasing carbon-limitation (Fig. 3B) and eventual cessation of growth (Fig. 3A).

@@

![Hysteresis under different nutrient and feeding regimes. Steady state S:H biomass ratios in constant environments with (A) low DIN (1e-7 M) and low food (2e-7 M), (B) low DIN (1e-7 M) and no food, (C) high DIN (2e-6 M) and low food (2e-7 M), and (D) high DIN (2e-6 M) and high food (4e-7 M). In each panel, steady states are shown starting from both high initial S:H biomass (1, i.e. healthy corals - closed circles) and low initial initial S:H biomass (0.0001, i.e. bleached corals - open circles). Points are colored according to whether the host exhibits positive (black) or negative (red) growth at steady state. Arrows above points indicate values beyond the y-axis range.](../img/Fig8.png)

@@

These dynamics reveal that the re-establishment of nitrogen-limitation is the most important diagnostic of recovery to a 'healthy' state, as this is when positive growth rates are resumed. A high S:H biomass ratio alone does not necessarily indicate that a 'healthy' state has been reached, since the carbon-limited state may still persist (e.g. Fig. 8B, 8C). This could explain why corals that have recovered their symbiont populations after bleaching may still exhibit energetic deficits and physiological impacts, possibly for months to years [@Levitan:2014et, @Hughes:2013hd]). These findings suggest that host acquisition of carbon from a source other than the symbiont may be extremely important for the system to recover from bleaching. Indeed, host feeding has been shown to promote a more rapid return to pre-bleaching levels of key physiological parameters in recovering corals [@Connolly:2012ep]. Additional carbon sources for the host, such as direct uptake of dissolved organic carbon [@Levas:2015ic], may also promote more rapid recovery from a bleached state.

>>>>

1. _Develop a new module of photo-oxidative stress._ Of primary interest to coral biologists and ecologists is symbiosis dysfunction under environmental stress, resulting in coral “bleaching”--the loss of algal symbionts from the association [@Jokiel:1977p7353]. Bleaching is thought to be triggered by photo-oxidative stress in *Symbiodinium* [@Weis:2008p944], which has been modeled previously as a response to absolute external irradiance [@Eynaud:2011tv]. However, this response may also depend on self-shading by other symbionts [@Enriquez:2005p142], CO~2~ availability at the site of photosynthesis [@Wooldridge:2009p7807], and other non-photochemical quenching mechanisms [@Roth:2014wf]. We incorporate these features into a new photo-oxidative stress module linking overreduction of the photosynthetic light reactions to downstream impacts of photoinhibition and photodamage. Importantly, this formulation introduces a link between CO~2~-limitation of photosynthesis and bleaching, and potential synergistic roles for heterotrophy and nutrient availability in influencing bleaching responses [@Wooldridge:2014hc].

@@

3. _Provide well-documented, open source code._ In order to facilitate the continued development and application of theoretical modeling tools for coral symbioses, we provide access to the model in the form of an R package called coRal ([github.com/jrcunning/coRal](github.com/jrcunning/coRal)), which users may install to run and visualize model simulations. With an accessible and modular framework, we envision this code base as a resource for further development by the scientific community to include additional complexity and problem-specific components. We chose R [@RALanguageandEn:2014wf], an open source programming language in common use by biologists and ecologists, to reach the widest possible audience with this work.

@@

Coral biomass formation is then specified by a parallel complementary SU [formula in Fig. 3.7 of @Kooijman:2010vd]. The general form for a production flux is $(m^{-1} + x^{-1} + y^{-1} - (x + y)^{-1})^{-1}$, where $m$ is the maximum production flux, and $x$ and $y$ are the input fluxes of the two substrates. Coral biomass is thus produced from carbon and nitrogen according to:

@@

where $\rho_C$ is fixed carbon shared by the symbiont (see Eq. 21), and $r_{NH}$ is recycled nitrogen liberated by host biomass turnover (see Eq. 7). The parameter $y_C$ specifies the yield of biomass from organic carbon, which we take to be 0.8 to satisfy redox balance [see @Muller:2009io].

@@

where $j_{CPm}$ is the maximum specific rate of photosynthesis, and $c_{ROS}$ is the relative rate ($\ge1$) of reactive oxygen species production (see Eq. 18). Dividing the photosynthetic rate by $c_{ROS}$ causes a decline in response to photooxidative stress at high light levels, and the emergent outcome of this SU formulation demonstrates a classic photoinhibition response (Fig. S2).

@@

The model dynamics are specified by the differential equations (1-2) that impose biomass balance for host and symbiont and by a set of coupled non-linear algebraic equations (3-22) that define fluxes. Several of these fluxes are defined *implicitly*; for example, the rejection fluxes of carbon and nitrogen from the symbiont and host biomass SUs, respectively, act as reciprocal input fluxes to the other SU. Similarly, the photosynthesis SU receives CO~2~ at a rate proportional to the carbon rejection flux from the host biomass SU, and the rejection flux of excitation energy from the photosynthesis SU acts to reduce its own production through photoinhibition. Without further assumptions, however, the dynamical system is not always unambiguously defined because for some combinations of parameters and environmental forcing functions the system of algebraic equations has more than one solution with all fluxes non-negative (see results below). In such circumstances, the right hand side of the differential equations (1) and (2) is not uniquely defined even when $S$ and $H$ are specified. We resolved this problem by *defining the dynamical system* as the limit as a time step $\Delta t \rightarrow 0$ of a discretized system corresponding to Euler integration of the differential equations, with those fluxes that represent flows of elemental matter implemented by assuming that transfer of material between components of the system takes one time step. Thus, for example, carbon rejected from the symbiont SU at time $t$ arrives at the host SU at time $t + \Delta t$.

Simulations using the discretized scheme were performed using R code developed in the coRal R package ([github.com/jrcunning/coRal](github.com/jrcunning/coRal)). By experimentation, we found that a time step of 0.1 days gave adequate precision for most simulations (including used to generate Figs. 2-8 in this paper). For steady state estimations, simulations were run until the changes in specific growth rate of the host and the S:H biomass ratio were less than 1e-5 per time step. In regions of state space where very slow transient dynamics could be expected (i.e. near bifurcation points), sample steady state calculations were verified using MATHEMATICA code for numerical root finding (function FindRoot) with the code written independently by a coauthor without reference to the R code. All of the R code for the simulations and figures presented in this paper can be found in the accompanying data repository at [github.com/jrcunning/coRal-analysis](github.com/jrcunning/coRal-analysis).

@@

At irradiance levels above ~25 mol photons m^-2^ d^-1^, steady state growth rates decline until positive growth ceases above ~40 µmol photons m^-2^ d^-1^ (Fig. 3A). The mechanism underlying this decline is the increase in light energy beyond the capacities of photosynthesis and non-photochemical quenching: excess excitation energy generates reactive oxygen species (ROS) [@Weis:2008p944; @Roth:2014wf], which, in this model, have the phenomenological consequences of reducing the photosynthetic rate (representing photoinhibition) and increasing symbiont biomass loss (representing photodamage and/or symbiont expulsion) [see @Eynaud:2011tv]. Together, these impacts reduce the symbiont to host biomass ratio (Fig. 3B), as occurs during coral bleaching. This reduction in symbionts consequently reduces the flux of fixed carbon to the host, resulting in increasing carbon-limitation and eventual cessation of growth (Fig. 3A).

@@

![Hysteresis under different nutrient and feeding regimes. Steady state S:H biomass ratios in constant environments with (A) low DIN (1e-7 mol N L^-1^) and low food (2e-7 C-molX L^-1^), (B) low DIN (1e-7 mol N L^-1^) and no food, (C) high DIN (2e-6 mol N L^-1^) and low food (2e-7 C-molX L^-1^), and (D) high DIN (2e-6 mol N L^-1^) and high food (4e-7 C-molX L^-1^). In each panel, steady states are shown starting from both high initial S:H biomass (1, i.e. healthy corals - closed circles) and low initial initial S:H biomass (0.0001, i.e. bleached corals - open circles). Points are colored according to whether the host exhibits positive (black) or negative (red) growth at steady state. Arrows above points indicate values beyond the y-axis range.](../img/Fig8.png)

@@

These dynamics reveal that the re-establishment of nitrogen-limitation is the most important diagnostic of recovery to a 'healthy' state, as this is when positive growth rates are resumed. A high S:H biomass ratio alone does not necessarily indicate that a 'healthy' state has been reached, since the carbon-limited state may still persist (e.g. Fig. 9B,C). This could explain why corals that have recovered their symbiont populations after bleaching may still exhibit energetic deficits and physiological impacts, possibly for months to years [@Levitan:2014et; @Hughes:2013hd]. These findings suggest that host acquisition of carbon from a source other than the symbiont may be extremely important for the system to recover from bleaching. Indeed, host feeding has been shown to promote a more rapid return to pre-bleaching levels of key physiological parameters in recovering corals [@Connolly:2012ep]. Additional carbon sources for the host, such as direct uptake of dissolved organic carbon [@Levas:2015ic], may also promote more rapid recovery from a bleached state.

### 2017-07-14 ms/ms.tex
text edits/corrections

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Develop a new module of photo-oxidative stress.} Of primary
  interest to coral biologists and ecologists is symbiosis dysfunction
  under environmental stress, resulting in coral ``bleaching''--the loss
  of algal symbionts from the association (Jokiel and Coles 1977).
  Bleaching is thought to be triggered by photo-oxidative stress in
  \emph{Symbiodinium} (Weis 2008), which has been modeled previously as
  a response to absolute external irradiance (Eynaud, Nisbet, and Muller
  2011). However, this response may also depend on self-shading by other
  symbionts (Enríquez, Méndez, and Iglesias-Prieto 2005),
  CO\textsubscript{2} availability at the site of photosynthesis
  (Wooldridge 2009), and other non-photochemical quenching mechanisms
  (Roth 2014). We incorporate these features into a new photo-oxidative
  stress module linking overreduction of the photosynthetic light
  reactions to downstream impacts of photoinhibition and photodamage.
  Importantly, this formulation introduces a link between
  CO\textsubscript{2}-limitation of photosynthesis and bleaching, and
  potential synergistic roles for heterotrophy and nutrient availability
  in influencing bleaching responses (Wooldridge 2014b)
\item
  \emph{Reduce theoretical and mathematical complexity.} Following the
  logic of Jager, Martin, and Zimmer (2013), we exclude certain features
  of formal DEB theory in order to capture behaviors of interest with
  the simplest possible formulation. Here, we present a model without
  reserves, maturity, or reproduction (see Kooijman 2010). This
  formulation precludes modeling the full life cycle of corals as
  reproduction, larval stages, and metamorphosis are not considered, but
  greatly reduces theoretical complexity and parameter numbers, which is
  advantageous given the relative paucity of data for corals. Moreover,
  dynamics of the symbiosis (i.e., changes in symbiont to host biomass,
  including bleaching and recovery) and coral biomass growth are
  efficiently captured with this simpler formulation, which also
  increases accessibility for biologists and ecologists without
  requiring expertise in DEB theory.
\item
  \emph{Provide well-documented, open source code.} In order to
  facilitate the continued development and application of theoretical
  modeling tools for coral symbioses, we provide access to the model in
  the form of an R package called coRal
  \url{github.com/jrcunning/coRal}, which users may install to run and
  visualize model simulations. With an accessible and modular framework,
  we envision this code base as a resource for further development by
  the scientific community to include additional complexity and
  problem-specific components. We chose R (R Core Team 2014), an open
  source programming language in common use by biologists and
  ecologists, to reach the widest possible audience with this work.
\end{enumerate}

@@

Coral biomass formation is then specified by a parallel complementary SU
(formula in Fig. 3.7 of Kooijman (2010){]}). The general form of this
equation is \((m^{-1} + x^{-1} + y^{-1} - (x + y)^{-1})^{-1}\), where
\(m\) is the maximum production flux, and \(x\) and \(y\) are the input
fluxes of the two substrates. Coral biomass is thus produced from carbon
and nitrogen according to:

@@

where \(\rho_C\) is fixed carbon shared by the symbiont (see Eq. 21),
and \(r_{NH}\) is recycled nitrogen liberated by host biomass turnover
(see Eq. 7). The parameter \(y_C\) specifies the yield of biomass from
organic carbon, which we take to be 0.8 to satisfy redox balance (see
Muller et al. (2009)).

@@

where \(j_{CPm}\) is the maximum specific rate of photosynthesis, and
\(c_{ROS}\) is the relative rate of reactive oxygen species production
(see Eq. 18). Dividing the photosynthetic rate by \(c_{ROS}\) causes a
decline in response to photooxidative stress at high light levels, and
the emergent outcome of this SU formulation demonstrates a classic
photoinhibition response (Fig. S2).

@@

The model dynamics are specified by the differential equations (1-2)
that impose biomass balance for host and symbiont and by a set of
coupled non-linear algebraic equations (3-22) that define fluxes.
Several of these fluxes are defined \emph{implicitly}; for example, the
rejection fluxes of carbon and nitrogen from the symbiont and host
biomass SUs, respectively, act as reciprocal input fluxes to the other
SU. Similarly, the photosynthesis SU receives CO\textsubscript{2} at a
rate proportional to the carbon rejection flux from the host biomass SU,
and the rejection flux of excitation energy from the photosynthesis SU
acts to reduce its own production through photoinhibition. Without
further assumptions, however, the dynamical system is not always
unambiguously defined because for some combinations of parameters and
environmental forcing functions the system of algebraic equations has
more than one solution with all fluxes non-negative (see results below).
In such circumstances, the right hand side of the differential equations
(1) and (2) is not uniquely defined even when \(S\) and \(H\) are
specified. We resolved this problem by \emph{defining the dynamical
system} as the limit as a time step \(\Delta t \rightarrow 0\) of a
discretized system corresponding to Euler integration of the
differential equations, with those fluxes that represent flows of
elemental matter implemented by assuming that transfer of material
between components of the system takes one time step. Thus, for example,
CO\textsubscript{2} rejected from the host SU at time \(t\) arrives at
the photosynthesis SU at time \(t + \Delta t\).

Simulations using the discretized scheme were performed using R code
developed in the coRal R package \url{github.com/jrcunning/coRal}. By
experimentation, we found that a time step of 0.1 days gave adequate
precision for most simulations (including used to generate Figs. 2-8 in
this paper). For steady state estimations, simulations were run until
the changes in specific growth rate of the host and the S:H biomass
ratio were less than 1e-5 per time step. In regions of state space where
very slow transient dynamics could be expected (i.e.~near bifurcation
points), sample steady state calculations were verified using
MATHEMATICA code for numerical root finding (function FindRoot) with the
code written independently by a coauthor without reference to the R
code. All of the R code for the simulations and figures presented in
this paper can be found in the accompanying data repository at
\url{github.com/jrcunning/coRal-analysis}.

@@

At irradiance levels above \textasciitilde{}25 mol photons
m\textsuperscript{-2} d\textsuperscript{-1}, steady state growth rates
decline until positive growth ceases above \textasciitilde{}40 µmol
photons m\textsuperscript{-2} d\textsuperscript{-1} (Fig. 3A). The
mechanism underlying this decline is the increase in light energy beyond
the capacities of photosynthesis and non-photochemical quenching: excess
excitation energy generates reactive oxygen species (ROS) (Weis 2008;
Roth 2014), which, in this model, have the phenomenological consequences
of reducing the photosynthetic rate (representing photoinhibition) and
increasing symbiont biomass loss (representing photodamage and/or
symbiont expulsion) (see Eynaud, Nisbet, and Muller 2011). Together,
these impacts reduce the symbiont to host biomass ratio (Fig. 3B), as
occurs during coral bleaching. This reduction in symbionts consequently
reduces the flux of fixed carbon to the host, resulting in increasing
carbon-limitation (Fig. 3B) and eventual cessation of growth (Fig. 3A).

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig8.png}
\caption{Hysteresis under different nutrient and feeding regimes. Steady
state S:H biomass ratios in constant environments with (A) low DIN (1e-7
M) and low food (2e-7 M), (B) low DIN (1e-7 M) and no food, (C) high DIN
(2e-6 M) and low food (2e-7 M), and (D) high DIN (2e-6 M) and high food
(4e-7 M). In each panel, steady states are shown starting from both high
initial S:H biomass (1, i.e.~healthy corals - closed circles) and low
initial initial S:H biomass (0.0001, i.e.~bleached corals - open
circles). Points are colored according to whether the host exhibits
positive (black) or negative (red) growth at steady state. Arrows above
points indicate values beyond the y-axis range.}
\end{figure}

@@

These dynamics reveal that the re-establishment of nitrogen-limitation
is the most important diagnostic of recovery to a `healthy' state, as
this is when positive growth rates are resumed. A high S:H biomass ratio
alone does not necessarily indicate that a `healthy' state has been
reached, since the carbon-limited state may still persist (e.g.~Fig. 8B,
8C). This could explain why corals that have recovered their symbiont
populations after bleaching may still exhibit energetic deficits and
physiological impacts, possibly for months to years (Levitan et al.
2014, Hughes and Grottoli (2013))). These findings suggest that host
acquisition of carbon from a source other than the symbiont may be
extremely important for the system to recover from bleaching. Indeed,
host feeding has been shown to promote a more rapid return to
pre-bleaching levels of key physiological parameters in recovering
corals (Connolly, Lopez-Yglesias, and Anthony 2012). Additional carbon
sources for the host, such as direct uptake of dissolved organic carbon
(Levas et al. 2015), may also promote more rapid recovery from a
bleached state.

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Develop a new module of photo-oxidative stress.} Of primary
  interest to coral biologists and ecologists is symbiosis dysfunction
  under environmental stress, resulting in coral ``bleaching''--the loss
  of algal symbionts from the association (Jokiel and Coles 1977).
  Bleaching is thought to be triggered by photo-oxidative stress in
  \emph{Symbiodinium} (Weis 2008), which has been modeled previously as
  a response to absolute external irradiance (Eynaud, Nisbet, and Muller
  2011). However, this response may also depend on self-shading by other
  symbionts (Enríquez, Méndez, and Iglesias-Prieto 2005),
  CO\textsubscript{2} availability at the site of photosynthesis
  (Wooldridge 2009), and other non-photochemical quenching mechanisms
  (Roth 2014). We incorporate these features into a new photo-oxidative
  stress module linking overreduction of the photosynthetic light
  reactions to downstream impacts of photoinhibition and photodamage.
  Importantly, this formulation introduces a link between
  CO\textsubscript{2}-limitation of photosynthesis and bleaching, and
  potential synergistic roles for heterotrophy and nutrient availability
  in influencing bleaching responses (Wooldridge 2014b).
\item
  \emph{Reduce theoretical and mathematical complexity.} Following the
  logic of Jager, Martin, and Zimmer (2013), we exclude certain features
  of formal DEB theory in order to capture behaviors of interest with
  the simplest possible formulation. Here, we present a model without
  reserves, maturity, or reproduction (see Kooijman 2010). This
  formulation precludes modeling the full life cycle of corals as
  reproduction, larval stages, and metamorphosis are not considered, but
  greatly reduces theoretical complexity and parameter numbers, which is
  advantageous given the relative paucity of data for corals. Moreover,
  dynamics of the symbiosis (i.e., changes in symbiont to host biomass,
  including bleaching and recovery) and coral biomass growth are
  efficiently captured with this simpler formulation, which also
  increases accessibility for biologists and ecologists without
  requiring expertise in DEB theory.
\item
  \emph{Provide well-documented, open source code.} In order to
  facilitate the continued development and application of theoretical
  modeling tools for coral symbioses, we provide access to the model in
  the form of an R package called coRal
  (\url{github.com/jrcunning/coRal}), which users may install to run and
  visualize model simulations. With an accessible and modular framework,
  we envision this code base as a resource for further development by
  the scientific community to include additional complexity and
  problem-specific components. We chose R (R Core Team 2014), an open
  source programming language in common use by biologists and
  ecologists, to reach the widest possible audience with this work.
\end{enumerate}

@@

Coral biomass formation is then specified by a parallel complementary SU
(formula in Fig. 3.7 of Kooijman 2010). The general form for a
production flux is \((m^{-1} + x^{-1} + y^{-1} - (x + y)^{-1})^{-1}\),
where \(m\) is the maximum production flux, and \(x\) and \(y\) are the
input fluxes of the two substrates. Coral biomass is thus produced from
carbon and nitrogen according to:

@@

where \(\rho_C\) is fixed carbon shared by the symbiont (see Eq. 21),
and \(r_{NH}\) is recycled nitrogen liberated by host biomass turnover
(see Eq. 7). The parameter \(y_C\) specifies the yield of biomass from
organic carbon, which we take to be 0.8 to satisfy redox balance (see
Muller et al. 2009).

@@

where \(j_{CPm}\) is the maximum specific rate of photosynthesis, and
\(c_{ROS}\) is the relative rate (\(\ge1\)) of reactive oxygen species
production (see Eq. 18). Dividing the photosynthetic rate by \(c_{ROS}\)
causes a decline in response to photooxidative stress at high light
levels, and the emergent outcome of this SU formulation demonstrates a
classic photoinhibition response (Fig. S2).

@@

The model dynamics are specified by the differential equations (1-2)
that impose biomass balance for host and symbiont and by a set of
coupled non-linear algebraic equations (3-22) that define fluxes.
Several of these fluxes are defined \emph{implicitly}; for example, the
rejection fluxes of carbon and nitrogen from the symbiont and host
biomass SUs, respectively, act as reciprocal input fluxes to the other
SU. Similarly, the photosynthesis SU receives CO\textsubscript{2} at a
rate proportional to the carbon rejection flux from the host biomass SU,
and the rejection flux of excitation energy from the photosynthesis SU
acts to reduce its own production through photoinhibition. Without
further assumptions, however, the dynamical system is not always
unambiguously defined because for some combinations of parameters and
environmental forcing functions the system of algebraic equations has
more than one solution with all fluxes non-negative (see results below).
In such circumstances, the right hand side of the differential equations
(1) and (2) is not uniquely defined even when \(S\) and \(H\) are
specified. We resolved this problem by \emph{defining the dynamical
system} as the limit as a time step \(\Delta t \rightarrow 0\) of a
discretized system corresponding to Euler integration of the
differential equations, with those fluxes that represent flows of
elemental matter implemented by assuming that transfer of material
between components of the system takes one time step. Thus, for example,
carbon rejected from the symbiont SU at time \(t\) arrives at the host
SU at time \(t + \Delta t\).

Simulations using the discretized scheme were performed using R code
developed in the coRal R package (\url{github.com/jrcunning/coRal}). By
experimentation, we found that a time step of 0.1 days gave adequate
precision for most simulations (including used to generate Figs. 2-8 in
this paper). For steady state estimations, simulations were run until
the changes in specific growth rate of the host and the S:H biomass
ratio were less than 1e-5 per time step. In regions of state space where
very slow transient dynamics could be expected (i.e.~near bifurcation
points), sample steady state calculations were verified using
MATHEMATICA code for numerical root finding (function FindRoot) with the
code written independently by a coauthor without reference to the R
code. All of the R code for the simulations and figures presented in
this paper can be found in the accompanying data repository at
\url{github.com/jrcunning/coRal-analysis}.

@@

At irradiance levels above \textasciitilde{}25 mol photons
m\textsuperscript{-2} d\textsuperscript{-1}, steady state growth rates
decline until positive growth ceases above \textasciitilde{}40 µmol
photons m\textsuperscript{-2} d\textsuperscript{-1} (Fig. 3A). The
mechanism underlying this decline is the increase in light energy beyond
the capacities of photosynthesis and non-photochemical quenching: excess
excitation energy generates reactive oxygen species (ROS) (Weis 2008;
Roth 2014), which, in this model, have the phenomenological consequences
of reducing the photosynthetic rate (representing photoinhibition) and
increasing symbiont biomass loss (representing photodamage and/or
symbiont expulsion) (see Eynaud, Nisbet, and Muller 2011). Together,
these impacts reduce the symbiont to host biomass ratio (Fig. 3B), as
occurs during coral bleaching. This reduction in symbionts consequently
reduces the flux of fixed carbon to the host, resulting in increasing
carbon-limitation and eventual cessation of growth (Fig. 3A).

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig8.png}
\caption{Hysteresis under different nutrient and feeding regimes. Steady
state S:H biomass ratios in constant environments with (A) low DIN (1e-7
mol N L\textsuperscript{-1}) and low food (2e-7 C-molX
L\textsuperscript{-1}), (B) low DIN (1e-7 mol N L\textsuperscript{-1})
and no food, (C) high DIN (2e-6 mol N L\textsuperscript{-1}) and low
food (2e-7 C-molX L\textsuperscript{-1}), and (D) high DIN (2e-6 mol N
L\textsuperscript{-1}) and high food (4e-7 C-molX
L\textsuperscript{-1}). In each panel, steady states are shown starting
from both high initial S:H biomass (1, i.e.~healthy corals - closed
circles) and low initial initial S:H biomass (0.0001, i.e.~bleached
corals - open circles). Points are colored according to whether the host
exhibits positive (black) or negative (red) growth at steady state.
Arrows above points indicate values beyond the y-axis range.}
\end{figure}

@@

These dynamics reveal that the re-establishment of nitrogen-limitation
is the most important diagnostic of recovery to a `healthy' state, as
this is when positive growth rates are resumed. A high S:H biomass ratio
alone does not necessarily indicate that a `healthy' state has been
reached, since the carbon-limited state may still persist (e.g.~Fig.
9B,C). This could explain why corals that have recovered their symbiont
populations after bleaching may still exhibit energetic deficits and
physiological impacts, possibly for months to years (Levitan et al.
2014; Hughes and Grottoli 2013). These findings suggest that host
acquisition of carbon from a source other than the symbiont may be
extremely important for the system to recover from bleaching. Indeed,
host feeding has been shown to promote a more rapid return to
pre-bleaching levels of key physiological parameters in recovering
corals (Connolly, Lopez-Yglesias, and Anthony 2012). Additional carbon
sources for the host, such as direct uptake of dissolved organic carbon
(Levas et al. 2015), may also promote more rapid recovery from a
bleached state.
